Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant 098051.}

### 2016-07-18 paper/paper.tex
Updated the experiments.

GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. The index uses fast succinct data structures instead of the smaller but slower compressed data structures in most of its components. The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0. All files were stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the queries.

@@

\begin{table}[t!]
\begin{center}
\caption{GCSA2 construction with different orders of the path graph. Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
\begin{tabular}{c|ccc|cc}
\hline
\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
\hline
 32 &        &         &        &         &         \\
 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[t!]
\begin{center}
\caption{Index sizes with different orders $k$ of the path graph. Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Section \ref{sect:extensions}.}\label{table:indexes}
\begin{tabular}{c|cc|c|c|c}
\hline
\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
\hline
 32 & 6.20G & & & & \\
 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
128 &  116G & 5.76G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
\hline
\end{tabular}
\end{center}
\end{table}

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index. Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cccc|cccc}
\hline
 & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
GCSA: 16 & 351584 & 347453 & 1829M &  872M & 18.0 \textmu{}s & 0.40 \textmu{}s & 0.93 \textmu{}s & 8.61 \textmu{}s \\
GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.7 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 75.3 \textmu{}s & 0.25 \textmu{}s & 0.33 \textmu{}s & 1.66 \textmu{}s \\
\hline
FM:   16 & 351584 & 301538 &    -- &  782M & 6.04 \textmu{}s &              -- &              -- & 2.44 \textmu{}s \\
FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.17 \textmu{}s \\
FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.92 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes, which is less than 1~bit per \kmer{k} for the \orderk{128} index. Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.

@@

\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques as GCSA2. While GCSA2 can index any graph but restricts the length of the queries, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.

\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2. The current implementation can only handle local variation.

>>>>

GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components. The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0. All files were stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.

@@

\begin{table}[t!]
\begin{center}
\caption{GCSA2 construction with different orders of the path graph. Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
\begin{tabular}{c|ccc|cc}
\hline
\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
\hline
 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[t!]
\begin{center}
\caption{Index sizes with different orders $k$ of the path graph. Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Section \ref{sect:extensions}.}\label{table:indexes}
\begin{tabular}{c|cc|c|c|c}
\hline
\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
\hline
 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
128 &  116G & 5.76G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
\hline
\end{tabular}
\end{center}
\end{table}

We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread. We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{128} GCSA and an FM\nobreakdash-index. Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cccc|cccc}
\hline
 & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
GCSA:  16 & 351584 & 347453 & 2477M &  872M & 18.3 \textmu{}s & 0.41 \textmu{}s & 0.87 \textmu{}s & 12.1 \textmu{}s \\
GCSA:  32 & 351555 & 333258 &  112M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 11.1 \textmu{}s \\
GCSA:  64 & 351567 & 326101 & 2.63M & 1.35M & 76.1 \textmu{}s & 0.26 \textmu{}s & 0.28 \textmu{}s & 5.76 \textmu{}s \\
GCSA: 128 & 351596 & 316500 & 0.32M & 0.37M &  149 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 5.98 \textmu{}s \\
\hline
FM:    16 & 351584 & 301538 &    -- &  782M & 6.06 \textmu{}s &              -- &              -- & 2.46 \textmu{}s \\
FM:    32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.19 \textmu{}s \\
FM:    64 & 351567 &  88184 &    -- & 0.84M & 17.2 \textmu{}s &              -- &              -- & 1.91 \textmu{}s \\
FM:   128 & 351596 &  35678 &    -- & 0.08M & 23.6 \textmu{}s &              -- &              -- & 3.48 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions. Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.

@@

\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} are based on the same techniques. While GCSA2 can index any graph but restricts query length, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.

\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2 would be. The current implementation can only handle local variation.
